---
title: "Kalshi vs Model: 12/10–12/19 Margin-of-Safety Buys"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
params:
  margin_of_safety: 0.10   # 10 percentage points
  stake_dollars: 10        # dollars per bet
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

theme_set(theme_minimal())

mos <- params$margin_of_safety
stake <- params$stake_dollars
```

## Data Sources

- Model predictions: `codex/glm_2025_future_games_1210_1220_predictions.csv`
- Kalshi snapshots: `codex/prices.csv`
- Kalshi code to sid map: `codex/kalshi_sid_map.csv`
- Realized winners: `2025/game_links.csv` (women)

```{r load}
# Paths (relative to this Rmd in codex/)
pred_path   <- "glm_2025_future_games_1210_1220_predictions.csv"
prices_path <- "prices.csv"
map_path    <- "kalshi_sid_map.csv"
links_path  <- file.path("..", "2025", "game_links.csv")

# Predictions (model win probabilities)
pred_raw <- read_csv(pred_path, show_col_types = FALSE) %>% clean_names()

pred <- pred_raw %>%
  mutate(
    game_date = as_date(game_date),
    team = as.character(team)
  ) %>%
  # Remove duplicates by team/date (keep first if any repeats exist)
  arrange(game_date, team) %>%
  distinct(game_date, team, .keep_all = TRUE)

# Kalshi price snapshots
prices_raw <- read_csv(prices_path, show_col_types = FALSE) %>% clean_names()

prices <- prices_raw %>%
  mutate(
    game_date_local = as_date(game_date_local),
    # Extract the final team code suffix from the ticker, e.g. "...-GONZ" -> "GONZ"
    kalshi_code = sub(".*-", "", canonical_ticker)
  ) %>%
  filter(status == "ok", !is.na(yes_ask_close_dollars))

# Mapping Kalshi code -> internal sid slug
code_map <- read_csv(map_path, show_col_types = FALSE) %>% clean_names() %>%
  select(kalshi_code, sid_guess)

# Realized winners (women only)
links_raw <- read_csv(links_path, show_col_types = FALSE) %>% clean_names()

links <- links_raw %>%
  filter(gender == "women") %>%
  mutate(
    # Convert mm/dd/yy -> ISO date
    game_date = as_date(mdy(game_date))
  ) %>%
  select(game_date, winner_sid, loser_sid)
```

## Join and Compute Edges

```{r join}
# Bring Kalshi codes to sid slugs
prices_sid <- prices %>%
  left_join(code_map, by = "kalshi_code") %>%
  rename(sid = sid_guess)

# Restrict date window to 2025-12-10 through 2025-12-19 inclusive
start_date <- as_date("2025-12-10")
end_date   <- as_date("2025-12-19")

prices_window <- prices_sid %>%
  filter(game_date_local >= start_date, game_date_local <= end_date)

# Join model probs to Kalshi snapshots by (date, team sid)
comp <- prices_window %>%
  inner_join(
    pred %>% select(game_date, team, glm_win_prob, team_name, opponent, opponent_name),
    by = c("game_date_local" = "game_date", "sid" = "team")
  ) %>%
  mutate(
    edge = glm_win_prob - yes_ask_close_dollars,
    mos_threshold = glm_win_prob - mos,
    buy_signal = yes_ask_close_dollars <= mos_threshold
  )

glimpse(comp)
```

```{r edge-summary}
edge_summary <- comp %>%
  summarise(
    n_markets = n(),
    mean_edge = mean(edge, na.rm = TRUE),
    median_edge = median(edge, na.rm = TRUE),
    pctl_90_edge = quantile(edge, 0.9, na.rm = TRUE)
  )

edge_summary
```

## Potential Buys (10% Margin)

```{r buys}
buys <- comp %>%
  filter(buy_signal) %>%
  arrange(desc(edge)) %>%
  select(
    game_date = game_date_local,
    team_sid = sid,
    team_name,
    opponent_sid = opponent,
    opponent_name,
    glm_win_prob,
    yes_ask_close_dollars,
    edge,
    canonical_ticker,
    title
  )

nrow(buys)
head(buys, 20)
```

## Realized Performance ($10 per buy)

Assumptions:

- Buy a YES position worth `$10` at the quoted ask.
- Payout per contract is $1 if the team wins, $0 if it loses.
- Fractional contracts allowed for simulation (contracts = stake / ask).
- Fees ignored for simplicity.

```{r pnl}
buys_scored <- buys %>%
  left_join(links, by = c("game_date" = "game_date")) %>%
  mutate(
    won = case_when(
      !is.na(winner_sid) & winner_sid == team_sid ~ TRUE,
      !is.na(loser_sid)  & loser_sid  == team_sid ~ FALSE,
      TRUE ~ NA
    ),
    contracts = stake / yes_ask_close_dollars,
    payout = case_when(
      won == TRUE  ~ contracts * 1.0,
      won == FALSE ~ 0.0,
      TRUE         ~ NA_real_
    ),
    profit = payout - stake,
    roi = profit / stake
  )

perf_summary <- buys_scored %>%
  summarise(
    bets = n(),
    decided = sum(!is.na(won)),
    wins = sum(won == TRUE, na.rm = TRUE),
    losses = sum(won == FALSE, na.rm = TRUE),
    hit_rate = wins / decided,
    total_staked = sum(stake, na.rm = TRUE),
    total_payout = sum(payout, na.rm = TRUE),
    net_profit = sum(profit, na.rm = TRUE),
    roi = net_profit / total_staked
  )

perf_summary
```

```{r show-buys}
buys_scored %>%
  arrange(desc(edge)) %>%
  select(game_date, team_sid, team_name, opponent_sid, opponent_name,
         glm_win_prob, yes_ask_close_dollars, edge, won, contracts, profit) %>%
  head(50)
```

## Notes

- Margin of safety is treated as absolute 10 percentage points (e.g., buy when ask <= model - 0.10). You can tweak via the `params$margin_of_safety` knob.
- The join uses `kalshi_sid_map.csv` to translate Kalshi codes to internal sids and matches on date + team sid.
- Markets with missing candles are excluded via `status == "ok"` and non-missing ask.
- Some prediction rows may list placeholder or duplicated `team_name`/`opponent_name`; the join keys use the sid slugs which are consistent across sources.
- Realized winners come from `2025/game_links.csv` and are filtered to `women`.

```{r session, echo=FALSE}
sessionInfo()
```
